---
title: "Stats 506, F20, Problem Set X"
author: "Your Name, email@umich.edu"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
     
---

```{r setup, include=FALSE}
# 79: -------------------------------------------------------------------------
#! We generally don't need to see the code in the results document.
knitr::opts_chunk$set(echo = FALSE)
#! Make sure all chunks have a label. This one is labelled 'setup'.
```

## Question 1

Use a first level header for each question.  

### Part A
Use lower level headers for thematic parts.

### Part B
Use numbered lists for specific parts of each question:

 i. Response to first item in part B.
    Maybe reference supporting tables and graphs.
 i. Response to second item in part B.

\pagebreak

## Question 2: Including Plots

Please start each question on a new page in pdf documents.

```{r q2_pressure_plot, echo=FALSE, fig.height=4, fig.cap=cap}
# Create a plot for the `pressure` data as in the default Rmd file: -----------
#! Use `paste` to construct captions while respecting the 79 char limit.
cap = paste0(
 "*Title.* All figures and tables should be numbered, titled ",
 "and have a descriptive caption."
)

plot(pressure)
```

There are `r nrow(pressure)` observations in the `pressure` data.

\pagebreak

## Question 3: Other considerations.

Start your answer to each question by describing the supporting files used to
produce the results to follow.

The three files below are used to answer question 3:
  - `0-psX_q3_data.sh` downloads the data
  - `1-psX_q3_clean.R` does some data cleaning and processing
  - `2-psX_q3_analysis.R` contains the statistical analysis.


```{r q3_prep, include=FALSE}
# Run the supporting scripts so this document to focus on reporting: ----------
#source('./1-psX_q3_clean.R')
#source('./2-psX_q3_analysis.R')
```





