---
title: "Problem Set 4, Solution"
author: "<a href='./index.html'>Statistics 506, Fall 2020</a>" 
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: journal
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse); library(data.table)
})
```

## About
This is an example solution to [Problem Set 4][1] for Stats 506 in Fall 2020. 

To build this document, run 
`Rscript -e "rmarkdown::render('./PS4_solution.Rmd')"` at the command line or
`bash ps4_make.sh` to build this document *after* running the solution scripts.

## Question 1 

In this quesiton, we use the 2009 and 2015 Residential Energy Consumption
Survey [RECS](https://www.eia.gov/consumption/residential/) data to profile the
quantities and types of televisions in US homes, by Census Region.

### Data Preparation

The script `0-ps4_q1_data.sh` downloads the source data files from the RECS
website. These files are then prepared for analysis in SAS by 
`1-ps4_q1_prepdata.sas`. Use the links below to view these files or refer to
the solutions at the Stats506_F20 GitHub repository. 

<details>
 <summary> `0-ps4_q1_data.sh` </summary>
 Click `Code` to view of `Hide` to collapse. 
```{r 0-ps4_q1_data.sh, comment = "", results = "asis"}
writeLines(c('```bash', readLines("./0-ps4_q1_data.sh"), '```'))
```
</details>


<details>
 <summary> `1-ps4_q1_prepdata.sas` </summary>
```{r 1-ps4_q1_prepdata.sas, comment = "", results = "asis"}
writeLines(c('```SAS', readLines("./1-ps4_q1_prepdata.sas"), '```'))
```
</details>

### Analysis

The analyses for both parts of this question are carried out in the script
`2-ps4_q1_analysis.sas` which can be viewed below. The script defines a 
number of SAS macros to facilitate the analysis. Briefly, these are:

  * `%recs_sum()` - a high level macro for summarizing a variable in the RECS
    data by group. 
  * `%est_mean()` - a macro used by `%recs_sum()` to specifically compute the
     mean of a variable by group. 
  * `%est_prop()` - a macro, like `%est_mean()`, used by `%recs_sum()` to 
     specifically compute the proportions for levels of a categorical variable
     by group. 
  * `%csv_export()` - a macro for writing a SAS data set to csv.  This macro
     is also used in the solution to question 2. 

<details>
 <summary> `2-ps4_q1_analysis.sas` </summary>
```{r 2-ps4_q1_analysis.sas, comment = "", results = "asis"}
writeLines(c('```SAS', readLines("./2-ps4_q1_analysis.sas"), '```'))
```
</details>
  
### Part a

In part **a** we compare the average number of televisions (`TVCOLOR`) in US 
homes in 2009 and 2015 by Census Region. The data displayed below is read from
the file `n_tv.csv` written by `2-ps4_q1_analysis.sas`.

#### i - 2009 and 2015 estimates

```{r q1_a_i}
n_tv = fread('./n_tv.csv')

rbind(
  n_tv[, .(Year = "2009", region, est = est09, lwr = lwr09, upr = upr09)],
  n_tv[, .(Year = "2015", region, est = est15, lwr = lwr15, upr = upr15)]
) %>%
  ggplot( aes(x = est, y = region, color = Year) ) +
   geom_point() +
   geom_errorbarh( aes(xmin = lwr, xmax = upr) ) +
   xlab('Average # of TVs') +
   ylab('') + 
   theme_bw() +
   scale_color_manual(values = c('darkblue', 'orange'))
```

#### ii - 2015 less 2009 differences

```{r q1_a_ii, fig.cap=cap2}
cap2 = paste(
  "**Figure 2.** *Differences (2015 less 2009) in Average # of Televisions", 
  "by Census Region.*"
)
n_tv %>%
  ggplot( aes(x = diff, y = region) ) +
  geom_point() +
  geom_errorbarh( aes(xmin = lwr_diff, xmax = upr_diff) ) +
  theme_bw() +
  ylab('') +
  xlab('Difference in Average # of Televisions (2015 less 2009)') +
  xlim(c(-.4, 0))
```

#### iii - tables

```{r q1_a_iii}
tab1_cap = paste(
  "**Table 1.** *Average number of televisions in US homes in 2009 and 2015,",
  "by Census Region.*"
)
n_tv[, .(
  `Census Region` = region,
  `2009` = sprintf('%4.2f (%4.2f, %4.2f)', est09, lwr09, upr09),
  `2015` = sprintf('%4.2f (%4.2f, %4.2f)', est15, lwr15, upr15),
  `Difference` = sprintf('%4.2f (%4.2f, %4.2f)', diff, lwr_diff, upr_diff)
)] %>%
  knitr::kable(format = 'html', cap = tab1_cap, align = c('rccc')) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
```


### Part b
In part **b** we compare the proportion of primary televisions by display type
for most used television (`TVTYPE1`) in US homes in 2009 and 2015 by 
Census Region. The data displayed below is read from
the file `tv_type.csv` written by `2-ps4_q1_analysis.sas`.

#### i - 2009 and 2015 estimates

```{r q1_b_i, fig.cap = cap3}
cap3 = paste(
  "**Figure 3.** *Proportion of primary television type, by",
  "Census Region.* "             
)

# read in tv type results
tv_type = fread('./type_tv.csv')

# make tv_type a factor based on total change
tt_levels = tv_type[, .(dbar = mean(diff)), .(tv_type)][order(-dbar), tv_type]
tv_type[, tv_type := factor(tv_type, ..tt_levels)]
rbind(
  tv_type[, .(Year = "2009", region, tv_type, 
              est = est09 * 100, lwr = lwr09 * 100, upr = upr09 * 100)],
  tv_type[, .(Year = "2015", region, tv_type, 
              est = est15 * 100, lwr = lwr15 * 100, upr = upr15 * 100)]
) %>%
  ggplot( aes(x = est, y = region, color = Year) ) +
   geom_point() +
   geom_errorbarh( aes(xmin = lwr, xmax = upr) ) +
   facet_wrap(~tv_type) + 
   xlab('% of primary TVs') +
   ylab('') + 
   theme_bw() +
   scale_color_manual(values = c('darkblue', 'orange'))
```

#### ii - 2015 less 2009 differences

```{r q1_b_ii, fig.cap=cap4}
cap4 = paste(
  "**Figure 4.** *Differences (2015 less 2009) in % of primary TVs for each", 
  "type by Census Region.*"
)
tv_type %>%
  ggplot( aes(x = diff * 100, y = region) ) +
  geom_point() +
  geom_errorbarh( aes(xmin = lwr_diff * 100, xmax = upr_diff * 100) ) +
  facet_wrap(~ tv_type) +
  geom_vline(xintercept = 0, lty = 'dashed') + 
  theme_bw() +
  ylab('') +
  xlab('Difference in Average # of Televisions (2015 less 2009)') 
```

#### iii - table

```{r q1_b_iii}
tab2_cap = paste(
  "**Table 2.** *Percent of primary televisions by type in US homes in 2009",
  "and 2015 for each Census Region.*"
)

tab2 = tv_type[order(tv_type, region), .(
  `Census Region` = region,
  `TV Type` = tv_type, 
  `2009` = sprintf('%3.1f <br>(%3.1f, %3.1f)', 
                   est09 * 100, lwr09 * 100, upr09 * 100),
  `2015` = sprintf('%3.1f <br>(%3.1f, %3.1f)', 
                   est15 * 100, lwr15 * 100, upr15 * 100),
  `Difference` = sprintf('%3.1f <br>(%3.1f, %3.1f)', 
                         diff * 100, lwr_diff * 100, upr_diff * 100)
)] %>%
  knitr::kable(
    format = 'html', cap = tab1_cap, align = c('rrccc'), escape = FALSE
  ) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)

for (i in 1:length(tt_levels)) {
  tab2 = kableExtra::group_rows(
    tab2, 
    group_label = tt_levels[i], 
    start_row = 1 + (i - 1) * 4, 
    end_row = 4 * i
  )
}
tab2
```

---

### Question 2 [25 points]

In this question we used the NHANES dentition and demographics 
data from PS3 in order to model the probability that a permanent tooth is 
present as a (non-linear) function of age and other demographics. I chose to 
focus on tooth `OHX03TC`, the upper right 1st molar. The solution uses logistic
regression, as required, and smoothing splines to allow the probability to vary
smoothly with age.  The full analysis can be viewed below.

<details>
 <summary> `3-ps4_q2_analysis.sas` </summary>
```{r 3-ps4_q2.sas, comment = "", results = "asis"}
writeLines(c('```SAS', readLines("./3-ps4_q2.sas"), '```'))
```
</details>

#### (a) Presense of a Permanent Tooth as a Function of Age

In this portion, we treated the data is iid.  For modeling purposes, age was
approximately centered at 33 years old and scaled to units of decades 
(10 years). Cubic smoothing splines for age were implemented with 13 knots
(interior break points) resulting in 16 basis functions. In addition,
demographic controls were included for gender, race, and college attendance 
(with or without graduation, and limited to those over 20). 

The relationship with age is shown in Figure 5.  See part c for a regression
table. 

```{r q2_afig, fig.cap = cap5}
cap5 = paste(
  "**Figure 5.** *Expected probability for presence of a permanent upper", 
  "right first molar.* This plot shows estimated probabilities at specific",
  "ages for the presence of a permanent upper right first molar with", 
  "(pointwise) 95% confidence intervals for these probabilities. The specific",
  "proabilities here are for a reference group of female gender, Mexican",
  "American race, and no college. The relationship with age is modeled as the",
  "same for all groups, so other groups would show the same shape with",
  "probabilities shifted up or down."
)
# predictions at specific ages
yhat_mod3 = fread('./yhat_mod3.csv')
ggplot(yhat_mod3, aes(x = 10 * age_c + 33, y = Predicted)) + 
  geom_point() + 
  geom_line() +
  geom_errorbar( aes(ymin = LCLM, ymax = UCLM) ) + 
  xlab('age') + 
  ylab('(modeled) probability that permanent tooth is present') + 
  theme_bw() 
```

#### (b) & (c) Comparison with Survey Weights

The model from part was refit using the medical examination survey weights and
the survey design. Weights were divided by the number of cohorts used, 4, so
that results represent an average estimated relationship over the time period
used. **Table 3** below presents a comparison of the survey results to the
results obtained in part **a**. 

```{r q2_b}
# model comparison table 
pars = fread('./pars.csv')
pars = pars[!is.na(mod3_se), ]

# make nicer labels
race = c(
  'race-1' = 'Mexian American', 
  'race-2' = 'Other Hispanic',
  'race-3' = 'Non-Hispanic White',
  'race-4' = 'Non-Hispanic Black', 
  'race-6' = 'Non-Hispanic Asian',
  'race-7' = 'Other' 
)
age = sprintf('f%i(age)', 1:16)
names(age) = sprintf('sm_age-%i', 1:16)

vars = c('Intercept' = 'Intercept', 
         'college' = 'College', 
         'gender-1' = 'Male', 
         race, 
         age)

pars[, 
 Variable := ifelse(is.na(ClassVal0), 
                     Variable, 
                     sprintf('%s-%i', Variable, ClassVal0) ) ]
pars[, `:=`(Variable = factor(vars[Variable], vars), ClassVal0 = NULL)]
pars = pars[order(Variable)]

## format as a table
ci = function(est, se) {
  bnds = matrix(
    rep(est, each = 3) + c(0, -qnorm(.975), qnorm(.975)) * rep(se, each = 3),
    3, length(est)
  )
  sprintf('%4.2f <br>(%4.2f, %4.2f)', bnds[1, ], bnds[2, ], bnds[3, ])
}

## caption
tab3_cap = paste(
  "**Table 3.** *Coefficients for Logistic Regression Models.* Coefficients",
  "are given on the logistic scale owing to the large coefficents for the",
  "the first two basis functions for age."
)

## the table
tab3 = pars[, .(Variable, 
         'Part a, Coefficient <br> (95% CI)' = ci(mod3_est, mod3_se), 
         'Part b, Coefficient <br> (95% CI)' = ci(mod4_est, mod4_se)
         )]

knitr::kable(
  tab3, format = 'html', escape = FALSE, caption = tab3_cap, align = c('rcc')
) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
```

Note that the large coefficients for the first two basis functions of age
could likely be reduced by including additional knots in the ~3-6 year age
range where this probability changes fastest. 

```{r q2_c}
tab4_cap = paste(
  "**Table 4.** *Comparison of predicted probabilities at specific ages.*",
  "The table shows predictions from the model without (a) and with (b)",
  "use of the survey weights and design. The specific",
  "proabilities here are for a reference group of female gender, Mexican",
  "American race, and no college."
)

# merge probabilities for table 4
yhat_mod4 = fread('./yhat_mod4.csv')

m3 = 
  yhat_mod3[, .(
    Age = 10 * age_c + 33,
    `Part a, p` = sprintf('%4.2f (%4.2f, %4.2f)', Predicted, LCLM, UCLM)
  )]

m4 = 
  yhat_mod4[, .(
    Age = 10 * age_c + 33,
    `Part b, p` = sprintf('%4.2f (%4.2f, %4.2f)', Predicted, LCLM, UCLM)
  )]

tab4 = merge(m3, m4, by = 'Age')
knitr::kable(tab4, format = 'html', caption = tab4_cap, align = 'rcc')  %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
```



[1]: <https://jbhender.github.io/Stats506/F20/PS4.html> "Problem Set 4"
