---
title: "Problem Set 3 Solution, <br>Stats 506, F20"
author: "James Henderson, PhD"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: united
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

## About
This is an example solution to Problem Set 3 for Stats 506 in Fall 2020. 

To build this document, run 
`Rscript -e "rmarkdown::render('./PS3_solution.Rmd')"` at the command line or
`bash ps3_make.sh` to build this document *after* running the scripts which
prepare the source data.

## Data preparation

The script `ps3_prep_data.do` merges the demographic and dentition data and
creates input data sets `demo_ps3.dta` and `ohxden_ps3.dta` used by the 
solutions that follow.  

<details>
 <summary> `ps3_prep_data.do` </summary>
```{r ps3_prep_data.do, comment = "", results = "asis"}
writeLines(c('```stata', readLines("./ps3_prep_data.do"), '```'))
```
</details>

<details>
 <summary> `ps3_prep_data.log` </summary>
```{r ps3_prep_data.log, comment = "", results = "asis"}
writeLines(c('```', readLines("./ps3_prep_data.log"), '```'))
```
</details>

## Question 1

In this question, we construct a balance table for missing or incomplete 
dentition exams using the 4 cohort NHANES data from the previous assignment.

The primary work for this question is done in `ps3_q1.do`. Click below to see
its contents or log. 
<details>
 <summary> `ps3_q1.do` </summary>
```{r ps3_q1.do, comment = "", results = "asis"}
writeLines(c('```stata', readLines("./ps3_q1.do"), '```'))
```
</details>

<details>
 <summary> `ps3_q1.log` </summary>
```{r ps3_q1.log, comment = "", results = "asis"}
writeLines(c('```', readLines("./ps3_q1.log"), '```'))
```
</details>

The script outputs csv files named according to the pattern 
`balance_table_*.csv`. These are assembled for presentation using R below.

```{r q1_balance_prep}
## formatting functions
pretty_p = function(p, min = 1e-3, fmt = '%5.3f') {
  out = ifelse(p < min, sprintf('p < %s', fmt), sprintf('p = %s', fmt))
  ifelse(p < min, sprintf(out, min), sprintf(out, p))
}
pwc = function(n) {
  format(n, big.mark = ',')
}

## read in balance tables
files = grep("^balance_table", dir(), value = TRUE)
tab_list = lapply(files, read.csv)
names(tab_list) = 
  str_replace_all(files, '.csv', '') %>% str_replace_all('balance_table_', '')


## format categorical variable tab
tab1 = bind_rows(
  mutate(tab_list[['1']], u20 = 1),
  mutate(tab_list[['0']], u20 = 0) 
)

## categorical variables
tab = tab1 %>%
  transmute(
    variable = variable,
    Subgroup = level, 
    `Complete` = sprintf('%s (%4.1f%%)', pwc(n_obs), pct_obs),
    `Incomplete/Missing` = sprintf('%s (%4.1f%%)', pwc(n_miss), pct_miss),
    p = ifelse(is.na(p), '--', pretty_p(p)),
    u20
  )

## format age tab
tab2 = tab_list[["age"]] %>%
  transmute(
    under_20,
    ohx = ifelse(ohx == 1, 'Complete', 'Incomplete/Missing'),
    disp = sprintf('%4.1f (%i, %i)', mean, lwr, upr),
    p = pretty_p(p)
  ) 

tab_age = tab2 %>%
  select(!"p") %>%
  pivot_wider(id_cols = 'under_20', 
              names_from = 'ohx',
              values_from = disp
            ) %>%
  left_join(
    {filter(tab2, !is.na(p)) %>% select(under_20, p)}, 
    by = 'under_20'
  ) %>%
  mutate(
    variable = 'all_age', # named so after all in table
    Subgroup = 'Age, mean (IQR)',
    u20 = under_20
  ) %>%
  select(variable, Subgroup, !"under_20")

## append as a single table and arrange for presentation
tab = bind_rows(tab, tab_age) %>%
  arrange(-u20, variable)
```

Here is the balance table.

```{r q1_balance}
cap1 = paste(
  "**Table 1.** *Association of select demographics with having a complete",
  "dention exam.* Only those who participated in the medical examination are",
  "included. "
)

tab_html = tab %>%
  select(!c("variable", "u20")) %>%
  knitr::kable(format = 'html', caption = cap1, align = 'r') %>%
  kableExtra::kable_styling("striped", full_width = TRUE)

## add grouping to rows

### individual vars
n_vars = tab %>%
    group_by(u20, variable) %>%
    summarize(n = n(), .groups = 'drop') %>%
    arrange(-u20)
n_cum = cumsum(n_vars$n)
grps = which(n_vars$n > 1)

for ( g in grps ) {
  label = sprintf('%s, n (%%)', str_to_title(n_vars$variable[g]))
  tab_html = tab_html %>%
    kableExtra::group_rows(label, n_cum[g - 1] + 1, n_cum[g])
}
### under 20
n_u20 = as.vector(with(tab, table(u20)))
tab_html = tab_html %>%
  kableExtra::group_rows("Age < 20 years", 1, n_u20[2]) %>%
  kableExtra::group_rows("Age $\\ge$  20 years", n_u20[2] + 1,  sum(n_u20))

tab_html
```

## Question 2

In this question, we use logistic regression to model the probability of a 
complete dentition exam as a function of the demographic variables.

The primary work for this question is done in `ps3_q2.do`. Click below to see
its contents or log. 
<details>
 <summary> `ps3_q2.do` </summary>
```{r ps3_q2.do, comment = "", results = "asis"}
writeLines(c('```stata', readLines("./ps3_q2.do"), '```'))
```
</details>

<details>
 <summary> `ps3_q2.log` </summary>
```{r ps3_q2.log, comment = "", results = "asis"}
writeLines(c('```', readLines("./ps3_q2.log"), '```'))
```
</details>

These scripts outputs the summary file `ps3_q2.xlsx` with tabs:
  * AIC - the models considered and their AIC's,
  * Top Model - regression output for the top model,
  * Top Model etable - alternative regression output. 
  
The final table is included here primarily as a demonstration. 

In the regression models, please note the following:
 - as *Table 1* suggests, there are no dentition exams for those under age 1;
   these participants have been excluded as they are missing structurally. 
 - to make the corresponding odds ratios easier to interpret, age has been
   centered and scaled to 10-year units. 

The code below reads and formats these results for display.  
```{r ps2_q2_format}
## AIC
aic_tab = readxl::read_xlsx('./ps3_q2.xlsx', sheet = 'AIC')
aic_tab = aic_tab %>%
  rename( `Stata Command` = model ) %>%
  mutate( model = paste0('m', 1:n()), delta = AIC - min(AIC)) %>%
  select(model, everything())

## regression table
term_map = c(
  "1.under_20" = "Under 20", 
  "2.gender" = "Female",
  "age_c" = "Age in decades, (centered)",
  "c.age_c#c.age_c" = "Age squared",
  "1.college" = "Some college or College graduate",
  "1.college#c.age_c" = "Age x College",
  "_cons" = "Reference Probability (Age 32/Male/No college)"
)
reg_tab = readxl::read_xlsx('./ps3_q2.xlsx', sheet = 'Top Model')
reg_tab = reg_tab %>%
  filter( !{grepl('b[.]', Term) | grepl('^o[.]', Term)}) %>%
  transmute(
    Covariate = term_map[Term],
    `OR (95% CI)` = 
      ifelse(Term == '_cons', 
       sprintf('%4.2f (%4.2f, %4.2f)', plogis(est), plogis(lwr), plogis(upr)),
       sprintf('%4.2f (%4.2f, %4.2f)', exp(est), exp(lwr), exp(upr))
      )
  )

```

```{r q2_regtab}
cap2 = paste(
  "**Table 2.** *Odds Ratios for a Complete Dentition Exam.*",
  "Overall the probability of a complete dentition exam is high. Those",
  "under 20, males, and college attendance were associated with higher",
  "probablity of a complete exam. Increasing age is associated with a lower",
  "probability of a complete exam for those under ~60 (see marginal effects",
  "in the log, not included here.)"
)
knitr::kable(reg_tab, caption = cap2) %>%
  kableExtra::kable_styling("striped")
```

```{r q3_aictab}
cap3 = paste(
  "**Table 3.** *AIC values for each logistic regression models considered.*",
  "Model 2 has the smallest AIC, but model 1 is likley preferred for",
  "simplicity as the improvement in model fit is small."
)
knitr::kable(aic_tab, caption = cap3) %>%
  kableExtra::kable_styling("striped")
```

## Question 3

In the final question, we create a visualization for a descriptive 
investigation of how tooth status is associated with age. 

The results to be plotted are in `ps3_q3_results.csv` and are produced by
`ps3_q3.do`.  You can view this script or its log below.

<details>
 <summary> `ps3_q3.do` </summary>
```{r ps3_q3.do, comment = "", results = "asis"}
writeLines(c('```stata', readLines("./ps3_q3.do"), '```'))
```
</details>

<details>
 <summary> `ps3_q3.log` </summary>
```{r ps3_q3.log, comment = "", results = "asis"}
writeLines(c('```', readLines("./ps3_q3.log"), '```'))
```
</details>

The results are read in using the code below.

```{r q3_reulsts, message = FALSE}
q3 = read_delim("ps3_q3_results.csv", delim = ',')

types = c(
  paste(c('3rd', '2nd', '1st'), 'molar'),
  paste(c('2nd', '1st'), 'bicuspid'),
  'cuspid',
  'lateral incisor',
  'central incisor'
)
types_all = c(types, rev(types), types, rev(types))
rl = c(rep('right', 8), rep('left', 16), rep('right', 8))

q3_clean = q3 %>%
  mutate(
    age = str_replace_all(
      str_split(age_grp, '-', 2, simplify = TRUE)[, 2], '\\)', ''),
    age = as.numeric(age),
    tooth_num = as.numeric(str_sub(tooth, 1, 2)),
    type = types_all[tooth_num],
    right_left = rl[tooth_num],
    upr_lwr = ifelse(tooth_num > 16, 'upper', 'lower'),
    tooth_name = paste(upr_lwr, right_left, type),
    Status = ohx_tc
  ) 
  
```


Figures 1a-c presents a visualization of this analysis, split up by tooth type.

### Figure 1 {.tabset .tabset-fade .tabset-pills}

#### Molars
```{r q3_figa, fig.cap = cap1a}
cap_template = paste(
  "**Figure %s.** *Tooth status as a function of age for %s.*",
  "Lines represents a linear interpolation of 6-year age buckets with the",
  "shaded areas providing (pointwise) 95%% confidence intervals."
)
cap1a = sprintf(cap_template, "1a", "molars")
q3_clean %>%
  filter(grepl('molar', tooth_name)) %>%
  ggplot(aes(x = age, y = p, group = Status, color = Status) )  +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = Status, color = NULL),
    alpha = 0.5
  ) + 
  geom_line() +
  facet_wrap(~tooth_name, nrow = 4) +
  theme_bw() +
  ylab('proportion of teeth')
```

#### Cuspids
```{r q3_figb, fig.cap = cap1b}
cap1b = sprintf(cap_template, "1b", "cuspids and bicuspids")
q3_clean %>%
  filter(grepl('cuspid', tooth_name)) %>%
  ggplot(aes(x = age, y = p, group = Status, color = Status) )  +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = Status, color = NULL),
    alpha = 0.5
  ) + 
  geom_line() +
  facet_wrap(~tooth_name, nrow = 4) +
  theme_bw() +
  ylab('% of teeth')
```

#### Incisors
```{r q3_figc, fig.cap = cap1c}
cap1c = sprintf(cap_template, "1c", "incisors")
q3_clean %>%
  filter(grepl('incisor', tooth_name)) %>%
  ggplot(aes(x = age, y = p, group = Status, color = Status) )  +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = Status, color = NULL),
    alpha = 0.5
  ) + 
  geom_line() +
  facet_wrap(~tooth_name, nrow = 4) +
  theme_bw() +
  ylab('% of teeth')
```
