---
title: "Problem Set 2 Solution, <br>Stats 506, F20"
author: "James Henderson, PhD"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: united
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About
This is an example solution to Problem Set 2 for Stats 506 in Fall 2020. 

To build this document, run 
`Rscript -e "rmarkdown::render('./PS2_solution.Rmd')"` at the command line or
`bash ps2_make.sh` to build this document *after* running the scripts which
prepare the source data.

## Question 1

### Code Organization

The solution to this question is split into four parts:

 - `0-ps1_q1_data.sh` uses the `wget` shell utility to download needed data.
 - `1-ps1_q1_prep_data.R` cleans and creates minimal data files for analysis.
    This file creates `ps2_q1.RData` containing 4 tibbles:
    * `recs09` - cleaned/minimal 2009 data
    * `recs15` - cleaned/minimal 2015 data
    * `w09` - replicate weights for the 2009 data
    * `w15` - replicate weights for the 2015 data.
 - `2-ps2_q1_analysis.R` - defines a function that can be used to estimate 
   either means for numeric or logical variables, and proportions for unique
   levels in factor or character variables. That function is then used to 
   produce the needed estimates, which are then combined to compute differences.
 - `PS2_Solution.Rmd` runs `2-ps2_q1_analysis.R` and builds the page you are 
   reading. Code for figures and tables can be found here.  
 - `3-ps2_make.sh` runs the first two data preparation scripts and then 
   builds this document. 
 
### Solution

```{r analysis, message = FALSE}
source('./2-ps2_q1_analysis.R')
```


#### Number of TVs {.tabset .tabset-fade .tabset-pills}

The figures and table below compare the average number of televisions per 
household in 2009 and 2015 by Census Division and rurality.  

##### Figures

```{r tv_n_annual_fig, fig.cap = cap1a}
cap1a = paste(
  "**Figure 1a.** *Average number of televisions per household.* Estimates use",
  "the 2009 and 2015 RECS data."
)

div_order = {
  tv_n %>%
   group_by(division) %>%
   summarize( tv_n = mean(tv_n), .groups = 'drop' ) %>%
   arrange(desc(tv_n)) 
}[['division']]

tv_n %>%
  mutate(
    division = factor(division, div_order), 
    `Urban Type` = rurality
  ) %>%
  ggplot(
    aes(x = division, y = tv_n, color = year)) +
  geom_point( position = position_dodge(.5) ) +
  geom_errorbar( 
    aes(ymin = tv_n_lwr, ymax = tv_n_upr),
    position = position_dodge(.5), width = 0.2 
  ) + 
  facet_wrap(~`Urban Type`) + 
  coord_flip() +
  theme_bw() +
  scale_color_manual(values = c('darkblue', 'darkred')) + 
  ylab('Average # of televisions') + 
  xlab('')
```

```{r tv_n_diff, fig.cap = cap1b}
cap1b = paste(
  "**Figure 1b.** *Difference in average number of televisions per household",
  "2015 less 2009.*"
)

div_order = {
  tv_n_diff %>%
   group_by(division) %>%
   summarize( diff = mean(diff), .groups = 'drop' ) %>%
   arrange(desc(diff)) 
}[['division']]

tv_n_diff %>%
  mutate(
    division = factor(division, div_order), 
    `Urban Type` = rurality
  ) %>%
  ggplot( aes(x = division, y = diff, color = `Urban Type`) ) +
  geom_hline(yintercept = 0, lty = 'dashed') + 
  geom_point( position = position_dodge(.5) ) +
  geom_errorbar( 
    aes(ymin = lwr, ymax = upr),
    position = position_dodge(.5), width = 0.2 
  ) + 
  coord_flip() +
  theme_bw() +
  scale_color_manual(values = c('darkred', 'darkblue')) + 
  ylab('Difference in average # of televisions (2015 less 2009)') + 
  xlab('') 
```

##### Table

```{r tv_n_table}
cap_tab1 = paste(
 "**Table 1.** *Average number of televisions per household 2009 and 2015.*",
 "Numbers in parantheses represent 95% confidence intervals."
)

tab1 = tv_n %>%
  mutate(
    pretty = sprintf('<div>%4.2f</div> <div>(%4.2f, %4.2f)</div>', 
                     tv_n, tv_n_lwr, tv_n_upr)
  ) %>%
  pivot_wider(
    id_cols = c('division', 'rurality'),
    names_from = 'year',
    values_from = 'pretty'
  )

tab1 = tab1 %>% 
  left_join(
    transmute(tv_n_diff, 
           division,
           rurality,
           Change =  sprintf('<div>%4.2f</div> <div>(%4.2f, %4.2f)</div>',
                             diff, lwr, upr) 
    ),
    by = c('division', 'rurality')
)

tab1 %>%
  select(
    `Census Division` = division, 
    `Urban Type` = rurality,
    `2009`,
    `2015`,
    Change
  ) %>%
  knitr::kable(
    format = 'html', 
    escape = FALSE, 
    align = 'llccc',
    cap = cap_tab1
  ) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
```

#### Type of Primary Television {.tabset .tabset-fade .tabset-pills}

##### Figures

```{r tv_type_annual, fig.cap = cap2a}
cap2a = paste(
  "**Figure 2a.** *Primary television type.* This figure shows the percent of",
  "homes with each television type for their most used television. Note the",
  "increase in the percent of homes with LED TV's and the corresponding",
  "decrease in standard tube televisions."
)
## order divisions
div_order = {
  tv_type %>%
   filter(tv_type == 'Standard Tube' & year == 2009) %>%
   group_by(division) %>%
   summarize( p_tube = mean(p), .groups = 'drop' ) %>%
   arrange(desc(p_tube)) 
}[['division']]

## order tv_type
type_order = {
  tv_type %>%
   filter(year == 2009) %>%
   group_by(tv_type) %>%
   summarize( p = mean(p), .groups = 'drop' ) %>%
   arrange(desc(p)) 
}[['tv_type']]

## construct a plot
tv_type %>%
  mutate(
    division = factor(division, div_order), 
    `Urban Type` = rurality,
    tv_type = factor(tv_type, type_order)
  ) %>%
  ggplot( aes(x = division, y = p, shape = year, color = `Urban Type`) ) +
  geom_point( position = position_dodge(.5) ) +
  geom_point( position = position_dodge(.5) ) +
  geom_errorbar( 
    aes(ymin = p_lwr, ymax = p_upr),
    position = position_dodge(.5), width = 0.2
  ) + 
  facet_wrap(~ tv_type) + 
  coord_flip() +
  theme_bw() +
  scale_color_manual( values = c('darkorange', 'black') ) + 
  ylab('% of primary televisions') + 
  xlab('') +
  ylim(c(0, 75))
```

```{r tv_type_diff, fig.cap = cap2b}
cap2b = paste(
  "**Figure 2b.** *Changes in primary television type, 2015 less 2009.*"
)

tv_type_diff %>%
  mutate(
    division = factor(division, div_order), 
    `Urban Type` = rurality,
    tv_type = factor(tv_type, type_order)
  ) %>%
  ggplot( aes(x = division, y = d, color = `Urban Type`) ) +
  geom_hline( yintercept = 0, lty = 'dashed', color = 'darkgrey' )  + 
  geom_point( position = position_dodge(.5) ) +
  geom_point( position = position_dodge(.5) ) +
  geom_errorbar( 
    aes(ymin = d_lwr, ymax = d_upr),
    position = position_dodge(.5), width = 0.2
  ) + 
  facet_wrap(~ tv_type) + 
  coord_flip() +
  theme_bw() +
  scale_color_manual( values = c('darkorange', 'black') ) + 
  ylab('change in % of primary televisions, 2015 less 2009') + 
  xlab('') 
```

##### Table

```{r tv_type_table}
cap_tab2 = paste(
 "**Table 2.** *Primary TV types in US households (%) in 2009 and 2015.*",
 "Numbers in parantheses are 95% confidence intervals."
)

tv_type = tv_type %>%
  mutate(
    pretty = sprintf('%4.1f<br>(%4.1f, %4.1f)', 
                     p, p_lwr, p_upr),
    tv_type = factor(tv_type, rev(type_order))
  )

## 2009/2015 by rurality
tab2_urb = tv_type %>%
  filter(rurality == 'Urban') %>%
  pivot_wider(
    id_cols = c('division', 'tv_type'),
    names_from = 'year',
    names_prefix = 'urban_',
    values_from = 'pretty',
    values_fill = '--'
  )
tab2_rul = tv_type %>%
  filter(rurality == 'Rural') %>%
  pivot_wider(
    id_cols = c('division', 'tv_type'),
    names_from = 'year',
    names_prefix =  'rural_',
    values_from = 'pretty',
    values_fill = '--'
  )
tab2 = left_join(tab2_urb, tab2_rul, by = c('division', 'tv_type'))

## changes
tab2_diff = tv_type_diff %>%
    transmute(
      division,
      rurality,
      tv_type = factor(tv_type, rev(type_order)),
      Change =  sprintf('%4.1f<br>(%4.1f, %4.1f)',
                        d, d_lwr, d_upr) 
    ) %>%
  pivot_wider(
    id_cols = c('division', 'tv_type'),
    names_from = 'rurality', 
    values_from = 'Change',
    values_fill = '--'
  )
tab2 = left_join(tab2, tab2_diff, by = c('division', 'tv_type') )

cn = c('Census Division', 'TV Type', 
       '2009', '2015', 'Change',
       '2009', '2015', 'Change'
       )
tab2 %>%
  select(
    division, 
    tv_type, 
    urban_2009, 
    urban_2015, 
    Urban,
    rural_2009,
    rural_2015,
    Rural
  ) %>%
  arrange(division, desc(tv_type)) %>%
  knitr::kable(
    format = 'html', 
    escape = FALSE, 
    align = 'llcccccc',
    col.names = cn,
    cap = cap_tab2
  ) %>%
  kableExtra::kable_styling("striped", full_width = TRUE) %>%
  kableExtra::add_header_above(header = c(' ' = 2, 'Urban' = 3, 'Rural' = 3))
```
