---
title: "Problem Set 1 Solution, Stats 506, F20"
author: "James Henderson, PhD"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: united
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About
This is an example solution to Problem Set 1 for Stats 506 in Fall 2020. 

To build this document, run 
`Rscript -e "rmarkdown::render('./PS1_solution.Rmd')"` at the command line.

To run all scripts in this assignment run `bash ./ps1_make.sh`. 

## Question 1
In question 1, you were asked to write shell scripts to download and prepare
data from the NHANES surveys. 

### Part a 
The script for part "a" is `ps1_q1_ohxden.sh`. This script first loops over 
the cohort identifiers and file name suffixes listed in 
`ps1_nhanes_files.txt` and downloads the oral health dentition examination
data for the specified years. We then use the included `cutnames.sh` utility
to identity the requested columns and, after verifying they all appear in 
the same order, extract the data in these columns and append into a single
multi-cohort file.

<details> 
<summary> Click to see the contents of `ps1_nhanes_files.txt`. </summary>
```{r q1_files, comment=""}
writeLines(readLines('./ps1_nhanes_files.txt'))
```
</details>

<details>
<summary> Click below to see the script `ps1_q1_ohxden.sh`.  
</summary>
```{r q1a, comment="", results="asis"}
 writeLines(c('```bash', readLines('./ps1_q1_ohxden.sh'), '```'))
```
</details>

The script produces an appended csv file with the following numbers of rows
and columns.

```{bash q1asize, comment = ""}
echo $(wc -l nhanes_ohxden.csv | cut -d' ' -f4) rows including header
echo $(< nhanes_ohxden.csv head -n1 | tr ',', ' ' | wc -w | tr '  ' ' ') columns
```

### Part b

In this part you were asked to modify the script from part "a" to download, 
extract, and append demographic data for the same set of cohorts. The only 
changes needed were to the file name and the regular expression used by 
`cutnames.sh` for extracting columns.  

<details>
<summary> Click below to see the script `ps1_q1_demo.sh`.  
</summary>
```{r q1b, comment="", results="asis"}
 writeLines(c('```bash', readLines('./ps1_q1_demo.sh'), '```'))
```
</details>

The script produces the file `nhanes_demo.csv` which has the following 
dimensions. 

```{bash q1bsize, comment = ""}
echo $(wc -l nhanes_demo.csv | cut -d' ' -f4) rows including header
echo $(< nhanes_demo.csv head -n1 | tr ',', ' ' | wc -w | tr '  ' ' ') columns
```

---

## Question 2

### Functions
In question 2 you were asked to write functions for creating, plotting, and
computing the area under ROC and PR curves. In the solution, you will see a 
function `.perf_counts()` which counts the true and false positives and 
negatives associated with each unique value of the predictor `yhat`.  This
function is used by `perf_roc()` and `perf_pr()`. 

The example solution for this question is in the file `ps1_q2.R`. 
<details>
<summary> Click to view `ps1_q2.R`. </summary>
```{r q2script, comment = "", results = "asis"}
writeLines(c('```R', readLines('./ps1_q2.R'), '```'))
```
</details>

### Results on the isolet data {.tabset .tabset-fade .tabset-pills}

```{r q2results, comment = "", message = FALSE}
# libraries: ------------------------------------------------------------------
library(tidyverse)

# functions: ------------------------------------------------------------------
source('./ps1_q2.R')

# isolet data: ----------------------------------------------------------------
path = './'
file = sprintf('%s/isolet_results.csv', path)
isolet = read.table(file, sep = ',', header = TRUE)

# areas under the curve: ------------------------------------------------------
auc = with(isolet, perf_roc(yhat, y, 'none'))$auc_roc
pr = with(isolet, perf_pr(yhat, y, 'none'))$auc_pr
```

For the `isolet_results` data with predictions `yhat` of whether `y` is a vowel
(1) or consonant (2), the areas under the curve are: 

  + AUC ROC `r sprintf('%4.2f', auc)`
  + AUC PR `r sprintf('%4.2f', pr)`. 

Use the buttons below to select the desired plot.

#### ROC ggplot

```{r q2_roc_ggplot, fig.cap = cap1}
cap1 = '**Figure 1 (ggplot2)**. *ROC curve for the isolet data.*'
with(isolet, perf_roc(yhat, y, 'ggplot2'))
```

#### ROC base

```{r q2_roc_base, fig.cap = cap2}
cap2 = '**Figure 1 (base).** *ROC curve for the isolet data.*'
with(isolet, perf_roc(yhat, y, 'base'))
```

#### PR ggplot

```{r q2_pr_ggplot2, fig.cap = cap3}
cap3 = '**Figure 2 (ggplot2).** *Precision-recall curve for the isolet data.*'
with(isolet, perf_pr(yhat, y, 'ggplot2'))
```

#### PR base

```{r q2_pr_base, fig.cap = cap4}
cap4 = '**Figure 2 (base).** *Precision-recall curve for the isolet data.*'
with(isolet, perf_pr(yhat, y, 'base'))
```

